Wealth Inequality in China: Evidence from the 2017 and 2019 CHFS

### 2025-04-24 report/draft.qmd
update

# Introduction
China has experienced remarkable economic growth in recent decades following its reform and opening-up policy initiated in 1978. However, this period has also been marked by a significant rise in income and wealth inequality [@piketty_capital_2019]. Some scholars have provided evidence suggesting that China's overall inequality peaked around 2010 and subsequently began to decline [@kanbur_great_2021; @zhang_survey_2021]. It is also well-documented that China's inequality is strongly intertwined with its urban-rural and coastal-inland divisions [@piketty_capital_2019; @zhang_survey_2021]. This paper utilizes data from the 2017 and 2019 China Household Finance Survey (CHFS) to examine the current state of China's wealth inequality. Specifically, it investigates whether the latest publicly available data supports the view that China's inequality has continued to decline since around 2010, as well as how China's wealth inequality relates to urban-rural and regional disparities.

@@

Consistent with the finding in @zhang_survey_2021 that China's inequality peaked in 2008 and 2009, @kanbur_great_2021 similarly document that China's inequality increased sharply after economic reforms but started to plateau around 2010. They find the national Gini coefficient peaked at about 0.525 in 2010, subsequently declining to approximately 0.476 in 2016 before rising slightly again. Consistent with Zhang (2021), they attribute this recent turnaround primarily to narrowing rural-urban and coastal-inland disparities, driven by demographic changes, tightening rural labor markets, and rising rural wages.

@@

The CHFS dataset has been extensively utilized in academic research. Studies leveraging the CHFS data have explored topics ranging from wealth distribution and inequality [@tan_earnings_2017], household finance in rural China [@gan_household_2014], household medical insurance expenditure [@li_household_2019], and many others.

::: {#tbl-tbl1}

@@

Notably, the growth rate of mean and median income from 2011 to 2017 is clearly higher than that of wealth but from 2017 to 2019, the growth of mean and median wealth surpasses that of income by a great margin.

@@

Financial equity notably increased from 2017 to 2019, with mean values rising substantially from 103,355 RMB to 161,199 RMB and median values more than doubling from 17,170 RMB to 39,251 RMB. Correspondingly, financial assets increased their share in total household wealth from around 12% in 2017 to approximately 15% in 2019. Land asset ownership, reported by about 44% of households in 2017 and 41% in 2019, closely aligns with China's rural population proportion of approximately 42% in 2017 and 40% in 2019 [@world_bank_rural_nodate]. Observed changes in land, commercial, and other assets should be interpreted cautiously, as it appears that some assets are redefined or newly added to the variables, which might partly reflect definitional adjustments rather than purely economic dynamics.

@@

@tbl-gini-decomposition-2017 and @tbl-gini-decomposition-2019 show the factor decomposition results for 2017 and 2019, respectively. Both years confirm housing assets' persistent and dominant role in driving China's wealth inequality, contributing over 72% to the overall inequality. This aligns with earlier studies: @knight_why_2022 reported that housing equity's contribution rose significantly from 64% in 2002 to 79% in 2013, while @tan_earnings_2017 documented housing's share at approximately 67% of household wealth in 2011. Financial equity emerged as another important factor, increasing its contribution from approximately 11% in 2017 to about 14% in 2019. In contrast, contributions from commercial assets declined notably, potentially due to definitional changes in the survey.

Overall, the structure of household wealth inequality in China remained relatively stable between 2017 and 2019. Housing assets continued to play a dominant role, with financial assets becoming increasingly significant, suggesting evolving household investment behaviors. The following part expands this analysis by conducting group-specific wealth inequality decompositions to understand the difference in wealth inequality across different rural and urban households and east and non-east regions.

@@

When benchmarked against the national decompositions for 2017 and 2019 (see \@tbl-gini-decomposition-2017 and \@tbl-gini-decomposition-2019), the urban–rural patterns largely reiterate the centrality of housing, land, and financial equity for Chinese wealth inequality.  What the residence‑specific results add, however, is a clearer sense of _relative_ importance: housing is even more dominant in cities than the country‑wide average, whereas land and business assets loom larger in rural portfolios than the aggregate figures would suggest.  At the same time, the contribution of financial assets in both residence types remains close to the national share, confirming that the rise of household financial investment is a broad‑based rather than location‑specific phenomenon.

@@

Relative to the country‑wide decompositions for 2017 and 2019, the regional results confirm the pervasive importance of housing, land and financial assets in shaping wealth inequality across China.  Yet they also reveal distinct regional configurations.  In the coastal East, the contribution of housing climbs to more than 80 per cent of total inequality—roughly eight percentage points above the national average—underscoring the outsized role of metropolitan real‑estate markets.  By contrast, in the Non‑East provinces housing accounts for just 58 per cent, while land and commercial (business) assets together explain nearly one‑quarter of inequality, double their combined share in the East.  Financial equity is a sizeable driver in both regions, but its influence is marginally greater in the interior (≈ 16 %) than along the coast (≈ 12 %), reflecting the recent expansion of retail investment beyond the major coastal cities.  Debt components continue to dampen inequality everywhere; however, the equalising effect of mortgage liabilities is slightly stronger in the East, where household borrowing is more closely tied to high‑value housing.  These contrasts suggest that policies aimed at moderating wealth disparities will need to account for regional heterogeneity—addressing speculative housing dynamics in the East while simultaneously fostering land‑market reforms and business development in the interior—within the common framework of wider financial inclusion.

@@

The summary statistics highlight significant and persistent disparities between urban and rural China. Urban households consistently exhibit higher mean and median incomes and wealth compared to rural households. Urban mean wealth was 3.48 times greater than rural mean wealth in 2017 and widened slightly to 3.78 times in 2019. Median urban wealth was 2.82 times higher than the rural median in 2017, though this ratio decreased slightly to 2.52 in 2019. Additionally, the Gini coefficient for wealth is consistently higher in rural areas, though both rural and urban Gini coefficients declined modestly between 2017 and 2019.

@@

The decomposition results indicate that about 55–57% of total wealth inequality is attributable to within-group disparities among urban and rural households. The between-group component, reflecting the urban-rural wealth divide, contributes approximately 33–34%. The overlap component, representing inequality due to intersecting distributions of urban and rural wealth, accounts for around 9–12% of total inequality. The consistency of these shares between 2017 and 2019 underscores the continued significance of the urban-rural divide in shaping China's wealth inequality.

@@

The above summary statistics shows that the mean and median wealth in the East is significantly higher than in the other regions in both 2017 and 2019. Notably, the mean and median wealth of the West-area households surpassed that of the Middle in 2019, but West's Gini coefficient rose to the highest among all regions in 2019 as well.

@@

The group decompsition shows that similar to the urban-rural decomposition, the between-group component contribues over 30% to the overall wealth inequality in both 2017 and 2019. The between group contribution saw a slight decrease from 2017 to 2019 with a similari increase in the residual component. This might signal some modest convergence in household wealth between regions.

>>>>

# Introduction
China's remarkable economic growth over the past five decades was initiated by the reform and opening-up policy launched in 1978, transitioning the country from a centrally planned to a market-oriented economy. From the openning up in 1978 to 2024, China's GDP growth rate averaged over 9% per year, and 800 million people were lifted out of poverty, with considerable improvment in the nation's health and education [@world_bank_world_2024]. This growth, however, has not been uniform across time, geography, or economic sectors. Coastal provinces such as Jiangsu and Guangdong experienced substantially higher growth rates, benefiting from early exposure to foreign investment and trade, while inland and rural areas lagged, leading to pronounced regional disparities [@liao_sixty_2016]. Sectorally, China's strong growth has been driven by fixed-capital investment and export-oriented manufacuring, which althoug effective in the past, start to show signs of both domestic and international imbalances [@world_bank_world_2024]. Since 2010, China's annual GDP growth rate has been declining from 10.6% in 2011 to 5.2% in 2023[@world_bank_gdp_2025].

Concurrently, China experienced a substantial rise in wealth and income inequality. The wealth share controlled by the top 10% increased markedly from approximately 40% in the late 1970s to nearly 70% by 2015, while the share held by the bottom half significantly decreased [@piketty_capital_2019]. Some studies indicate that China's overall inequality reached its peak around 2010, after which it started to gradually decline [@kanbur_great_2021; @zhang_survey_2021]. Additionally, it is well-established that China's inequality closely correlates with its urban-rural and coastal-inland divisions, with urban and coastal regions exhibiting substantially higher levels of income and wealth compared to their rural and inland counterparts [@piketty_capital_2019; @zhang_survey_2021].

This paper analyzes the contemporary landscape of wealth inequality in China using recent data from the 2017 and 2019 China Household Finance Survey (CHFS). It specifically investigates whether recent trends confirm claims of continued inequality reduction since 2010. Through factor decomposition, the study examines the sources of wealth inequality within the entire population as well as separately within urban-rural and coastal-inland subgroups, highlighting variations in inequality among these different groups. Furthermore, the study applies group decomposition methods to assess the extent to which rural-urban and coastal-inland disparities contribute to overall wealth inequality. Through these detailed decompositions and analyses, the paper aims to provide a comprehensive and nuanced understanding of current patterns and dynamics in China's wealth inequality.

@@

Consistent with the finding in @zhang_survey_2021 that China's inequality peaked in 2008 and 2009, @kanbur_great_2021 similarly document that China's inequality increased sharply after economic reforms but started to plateau around 2010. They find the national Gini coefficient peaked at about 0.525 in 2010, subsequently declining to approximately 0.476 in 2016 before rising slightly again. Consistent with @zhang_survey_2021, they attribute this recent turnaround primarily to narrowing rural-urban and coastal-inland disparities, driven by demographic changes, tightening rural labor markets, and rising rural wages.

@@

The CHFS dataset has been extensively utilized in academic research. Studies leveraging the CHFS data have explored topics ranging from wealth distribution and inequality [@tan_earnings_2017], household finance in rural China [@gan_report_2014], household medical insurance expenditure [@liu_health_2022], and many others.

# Analysis
## Summary Statistics
::: {#tbl-tbl1}

@@

Notably, while the growth rate of mean and median income from 2011 to 2017 substantially outpaced that of wealth, this pattern reversed between 2017 and 2019, where the growth of mean and median wealth greatly surpassed income growth. This later-period divergence may reflect factors such as significant asset price appreciation relative to income gains.

@@

Financial equity notably increased from 2017 to 2019, with mean values rising substantially from 103,355 RMB to 161,199 RMB and median values more than doubling from 17,170 RMB to 39,251 RMB. Correspondingly, financial assets increased their share in total household wealth from around 12% in 2017 to approximately 15% in 2019. Land asset ownership, reported by about 44% of households in 2017 and 41% in 2019, closely aligns with China's rural population proportion of approximately 42% in 2017 and 40% in 2019 [@world_bank_rural_nodate]. However, later results in this paper show that the gini coefficient for land in the rural area was 89.7% in the rural area in 2019, which is significantly higher than the gini coefficient for wealth and for housing in rural areas. This shows that dispite a wide distribution of land ownership, the distribution of land is highly unequal in modern China.

Observed changes in land, commercial, and other assets should be interpreted cautiously, as it appears that some assets are redefined or newly added to the variables, which might partly reflect definitional adjustments rather than purely economic dynamics.

@@

@tbl-gini-decomposition-2017 and @tbl-gini-decomposition-2019 show the factor decomposition results for 2017 and 2019, respectively. Both years confirm housing assets' persistent and dominant role in driving China's wealth inequality, contributing over 72% to the overall inequality (72.7% in 2017, 72.3% in 2019). This aligns with earlier studies: @knight_why_2022 reported that housing equity's contribution rose significantly from 64% in 2002 to 79% in 2013, while @tan_earnings_2017 documented housing's share at approximately 67% of household wealth in 2011. Financial equity emerged as another important factor, increasing its contribution from approximately 11.5% in 2017 to about 13.5% in 2019. In contrast, contributions from commercial assets declined notably, potentially due to definitional changes in the survey.

Overall, the structure of household wealth inequality in China remained relatively stable between 2017 and 2019. Housing assets continued to play a dominant role, with financial assets becoming increasingly significant, suggesting evolving household investment behaviors. To explore heterogeneity in these patterns, the following sections decompose wealth inequality within specific demographic subgroups.

@@

When benchmarked against the national decompositions for 2017 and 2019 (see @tbl-gini-decomposition-2017 and @tbl-gini-decomposition-2019), the urban–rural patterns largely reiterate the centrality of housing, land, and financial equity for Chinese wealth inequality.  What the residence‑specific results add, however, is a clearer sense of _relative_ importance: housing is even more dominant in cities than the country‑wide average, whereas land and business assets loom larger in rural portfolios than the aggregate figures would suggest.  At the same time, the contribution of financial assets in both residence types remains close to the national share, confirming that the rise of household financial investment is a broad‑based rather than location‑specific phenomenon.

@@

Relative to the country‑wide decompositions for 2017 and 2019, the regional results confirm the pervasive importance of housing, land and financial assets in shaping wealth inequality across China, yet they also reveal distinct regional configurations. In the coastal East, housing's influence is particularly pronounced; its contribution climbs to over 80 per cent (80.3%) of total inequality—roughly eight percentage points above the national average—underscoring the outsized role of metropolitan real‑estate markets and its strong alignment with overall wealth rank (Gini correlation 0.96). By contrast, in the Non‑East provinces housing accounts for a lower share at 58.3 per cent. Here, land and commercial (business) assets play a much larger role, together explaining nearly one-quarter of inequality (23.2%, combining 13.8% from land and 9.4% from commercial), substantially higher than their combined share of about 7.6% in the East. This reflects not only portfolio differences but also stronger correlations between these assets and wealth rank in the Non-East (Gini correlations of 0.83 for land and 0.80 for commercial, vs. 0.35 and 0.72 respectively in the East).

Financial equity is a sizeable driver in both regions, though its contribution and correlation with wealth rank are marginally greater in the interior (15.7% contribution, 0.78 correlation) than along the coast (12.4% contribution, 0.77 correlation), perhaps reflecting the recent expansion of retail investment beyond major coastal cities. Debt components, particularly mortgage liabilities, serve to dampen inequality in both regions through their negative contributions to the overall Gini coefficient. These contrasts underscore that policies aimed at moderating wealth disparities need to account for regional heterogeneity—potentially addressing speculative housing dynamics in the East while simultaneously fostering land‑market reforms and business development in the interior—within a common framework of wider financial inclusion.

@@

The summary statistics highlight significant and persistent disparities between urban and rural China. Urban households consistently exhibit higher mean incomes and wealth compared to rural households. Urban mean wealth was 3.48 times greater than rural mean wealth in 2017 and widened slightly to 3.78 times in 2019. Additionally, the Gini coefficient for wealth is consistently higher in rural areas (0.6945 in 2017, 0.6629 in 2019) compared to urban areas (0.6573 in 2017, 0.6165 in 2019), though Gini coefficients for both groups declined modestly between the two years.

@@

The decomposition results indicate that about 55–57% (55.4% in 2017, 56.8% in 2019) of total wealth inequality is attributable to within-group disparities among urban and rural households. The between-group component, reflecting the urban-rural wealth divide, contributes approximately 33–34% (32.6% in 2017, 33.7% in 2019). The overlap component, representing inequality due to intersecting distributions of urban and rural wealth, accounts for around 9–12% (12.0% in 2017, 9.5% in 2019) of total inequality. The consistency of these shares between 2017 and 2019 underscores the continued significance of the urban-rural divide in shaping China's wealth inequality.

@@

The above summary statistics shows that the mean and median wealth in the East is significantly higher than in the other regions in both 2017 and 2019. Notably, the mean wealth of West-area households surpassed that of the Middle region in 2019, although its median wealth remained lower. Concurrently, the West's Gini coefficient rose to become the highest among all regions in 2019 (0.6947).

@@

The group decomposition shows that, similar to the urban-rural decomposition, the between-group component contributes over 30% to overall wealth inequality in both 2017 and 2019 (36.9% and 33.8% respectively). The between-group contribution saw a slight decrease from 2017 to 2019, with a similar increase in the residual component (from 32.0% to 34.5%). This might signal some modest convergence in household wealth between regions.

## Conclusion
This paper analyzed wealth inequality in China using the 2017 and 2019 CHFS data, finding a modest decline in the overall Gini coefficient (0.701 to 0.667), consistent with literature suggesting a post-2010 peak [@kanbur_great_2021; @zhang_survey_2021]. Factor decomposition confirmed housing assets as the primary driver (contributing over 72%), aligning with previous studies [@knight_why_2022; @tan_earnings_2017], while financial equity's contribution grew (from approximately 11% to 14%). Partitioned analysis highlighted housing's stronger impact in urban areas (around 75% vs. approximately 52% rural) and especially the East region (over 80% vs. approximately 58% Non-East), with divergent trends emerging between 2017-2019 (increasing contribution in the East, decreasing elsewhere). Group decomposition further revealed that while within-group inequality dominates (accounting for approximately 55-60%), substantial inequality stems from persistent urban-rural (around 33%) and regional (over 30%) divides.

These dynamics likely reflect intertwined factors: rapid urbanization, uneven real estate appreciation and speculation impacting housing's role; financial market development driving the rise of financial assets, potentially favoring wealthier urban households; and deep-rooted structural issues like historical policies, the Hukou system, and differential access to opportunities maintaining the significant urban-rural and regional gaps, despite potential slow convergence hinted by recent trends and policies [@kanbur_great_2021]. Consequently, addressing China's high wealth inequality requires a nuanced approach beyond national measures, necessitating context-specific policies such as targeted housing market regulation in the East, support for land and business development in rural and Non-East areas, and broader efforts towards financial inclusion across the country.
